# Data handling and visualization

As noted earlier in this guide, an R workflow basically consists of performing 
calculations and/or visualizing your data which has previously been imported into 
R. In the majority of cases, this workflow usually includes an intermediary step 
referred to as data manipulation meant to get your data into shape for 
visualization. Note that some packages even require you to provide the data to 
be displayed in a particular format (of which more later on).  

Within the scope of this short course, we'll simply assume that you already 
heard a lot about importing your data into R. Even if this were not the case, 
there's plenty of tutorials on the internet demonstrating the use of `read.table` 
and the like. Help on this topic may be found e.g. on 
[Quick-R](http://www.statmethods.net/input/importingdata.html), 
[CRAN](https://cran.r-project.org/doc/manuals/r-release/R-data.html), 
[R-Tutor](http://www.r-tutor.com/r-introduction/data-frame/data-import) and so 
forth. Just remember, Google is your friend &#9786;

<hr>

<b>Data manipulation</b>

A hodgepodge of typical data manipulation steps including illustrative examples 
is given at [Cookbook for R](http://www.cookbook-r.com/Manipulating_data/) and 
includes, amongst others,

* _General operations_
    * Sorting
    * Converting between vector types
    * Dealing with duplicate records and missing values
* _Factor operations_
    * Renaming and re-computing factor levels
    * Reordering factor levels (which comes in quite handy when it comes to plotting)
* _Data frame operations_
    * Adding and removing columns
    * Reordering columns
    * Merging data frames
* _Restructuring data_
    * Data format conversions
    * Summarizing data

Assuming that you're already familiar with the basics of data manipulation, we'll 
primarily focus on the latter point which represents an essential step towards 
conveying your results clearly and vividly to a broader audience. 

<hr>

<b>The `diamonds` dataset</b>

In the following, I'll demonstrate the essentials of data restructuring in R on 
the basis of the `diamonds` dataset included in **ggplot2**. In case you're not 
already familiar with the dataset, just take a minute to have a brief look at 
what it's all about. Also, feel free to try out `?diamonds` in order to get a 
more detailed description of the single variables.

Now, `diamonds` is not a standard R `data.frame`, but first and foremost an object of class `tbl_df`.

```{r tbl_df1, message = FALSE}
library(ggplot2)

class(diamonds)
```

One good thing about such objects in R is that their `print()` method largely renders ritual-like `head()`, `tail()` or `str()` calls unnecessary. 

```{r tbl_df2}
diamonds
```

Notice that not only the class and the dimensions of the dataset are displayed, 
but also the variable types of the single columns. Furthermore, the console 
output is truncated after the first 10 rows (which would also apply to the 
displayed columns if we were dealing with a somewhat wider dataset). Practically, 
this means that the appropriate use of `tbl_df` lets you kill two birds with one 
stone compared with the 2-step approach via `head` and `str` that is usually required for investigating standard data frames. 


## Summarizing data

The **dplyr** package is a marvelous and very intuitive set of tools to perform 
fast and easy data manipulation and summarization in preparation e.g. for 
visualization. In the following, we'll guide you through a selection of 
operations you may find helpful within the course of future R workflows on the 
basis of the 'diamonds' dataset. Based on a very comprehensive 
[RPubs tutorial](https://rpubs.com/justmarkham/dplyr-tutorial), these include

* `filter`, 
* `select`, 
* `arrange`, 
* `mutate` and 
* `summarise` (incl. `group_by`) 

each of which will be accompanied by an equivalent base-R approach.

<hr>

<b>Subsetting data via `filter`</b>

Just like base-R `subset` or simply indexing via conditional expressions 
included in square brackets, `filter` creates a subset of your data based on 
some user-defined criteria. For the time being, let's have a look at your 
subsetting abilities.

<hr>

<b>Task: base-R subsetting</b>

Suppose we want to create a subset of 'diamonds' keeping all D-, E- or F-colored 
specimen with a 'Premium' or 'Ideal' cut quality and a weight of more than 3 
carat. Try to figure out two different solutions, one <u>with</u> and one 
<u>without</u> using `subset`. 

<center>
  <img src="https://upload.wikimedia.org/wikipedia/commons/2/25/Hourglass_2.svg" alt="hourglass" style="width: 125px;"/>
</center>

```{r subset, echo = FALSE, eval = FALSE}
subset(diamonds, carat > 3 & 
         (cut == "Premium" | cut == "Ideal") & 
         (color == "D" | color == "E" | color == "F"))

subset(diamonds, carat > 3 &
         cut %in% c("Premium", "Ideal") &
         color %in% c("D", "E", "F"))

diamonds[diamonds$carat > 3 & 
           diamonds$cut %in% c("Premium", "Ideal") &
           diamonds$color %in% c("D", "E", "F"), ]
```

Using `filter` is straightforward and very similar to `subset`. The first 
argument represents the data frame, whereas any subsequent argument represents a 
logical expression to filter particular rows. Note that in contrast to the 
direct subsetting via square brackets, you are not required to repeat the name 
of the data frame with every single condition.

```{r filter}
library(dplyr)

filter(diamonds, carat > 3 &
         cut %in% c("Premium", "Ideal") &
         color %in% c("D", "E", "F"))
```

<hr>

<b>Selecting columns via `select`</b>

Selecting specific columns from a data frame follows a similar syntax to 
`filter` and works analogous to SQL standards. Again, the first argument 
represents the data frame under consideration followed by the desired column 
names (without double quotes). 

```{r select}
dplyr::select(diamonds, carat, cut, color, clarity)
```

Note that both the use of `c` to combine the single column names into a vector 
as well as the need for double quotes became obsolete. In addition to this, 
column names may be treated analogous to numeric indices, eliminating the need 
to count columns when desiring to extract several consecutive columns from a 
rather wide datasets. 

```{r select2}
dplyr::select(diamonds, carat:clarity, price)
```

There's also a set of additional helper functions, including `starts_with`, 
`ends_with`, `matches` and `contains`, which definitely go beyond the scope of 
this short introduction. Still, it's good to know that such things existed in 
case you needed any of them.

<hr>

<b>A brief note on chaining (or pipelining)</b>

Now suppose we wanted to select the same columns from the previously created 
subset of data. Traditionally, this would either require a 2-step approach, 
generating otherwise unnecessary intermediate results, via 

```{r filter_select, eval = FALSE}
## first, create subset
diamonds_sub <- filter(diamonds, carat > 3 &
                         cut %in% c("Premium", "Ideal") &
                         color %in% c("D", "E", "F"))

## second, select columns
select(diamonds_sub, carat:clarity, price)
```

or nested function calls which are usually hard to read.

```{r filter_select2, eval = FALSE}
## all-in-one nested solution
select(
  filter(diamonds, carat > 3 &
           cut %in% c("Premium", "Ideal") &
           color %in% c("D", "E", "F")), 
  carat:clarity, price
)
```

**dplyr** introduces the `%>%` operator which is meant to bridge a set of 
connected processing steps, thus eliminating the need for intermediary 
variables ('diamonds_sub') or nested functions. Just think of `%>%` as a "then" 
connecting two parts of a sentence.

```{r pipelining}
diamonds %>%
  filter(carat > 3 &
           cut %in% c("Premium", "Ideal") &
           color %in% c("D", "E", "F")) %>%
  dplyr::select(carat:clarity, price)
```

Note that chaining comes in handy when performing multiple operations at once, 
rendering your code much more elegant and reducing the accumulated overhead 
significantly. The single worksteps can be read from left to right and from top 
to bottom, just like you would read the pages of a book. 

<hr>

<b>Reordering data frames via `arrange`</b>

`arrange` ensures a fast and easy rearrangement of rows based on certain 
variables. Although this can be done using base-R, its necessity soon becomes 
clear when dealing with rearrangements based on multiple variables as the **dplyr** 
approach requires far less typing.

<hr>

<b>Task: base-R rearrangement of rows</b>

Here's a tricky one. Let's assume we wanted to rearrange the rows of a subset of 
'diamonds' (color, price, carat) according to color, with the best color (D) on 
top. Since we're also interested in a preferably low price, the diamonds of 
uniform color should be sorted according to their price, with cheapest ones on 
top. Finally, for each set of diamonds with a specified color and price, we want 
their weights arranged in descending order, with the heaviest specimen on top.

<center>
  <img src="https://upload.wikimedia.org/wikipedia/commons/2/25/Hourglass_2.svg" alt="hourglass" style="width: 125px;"/>
</center>

```{r order, echo = FALSE, eval = FALSE}
diamonds_sub <- diamonds[, c("color", "price", "carat")]
diamonds_carat <- diamonds_sub[order(diamonds_sub$carat, decreasing = TRUE), ]
diamonds_carat[order(diamonds_carat$color, diamonds_carat$price), ]
```

Again, the **dplyr** approach is much simpler and can be done in one go. 

```{r arrange}
diamonds %>%
  dplyr::select(color, price, carat) %>%
  arrange(color, price, desc(carat))
```

<hr>

<b>Add new columns via `mutate`</b>

`mutate` lets you add new variables to an existing data frame. It is basically 
an extended version of base-R `transform` in the sense that it allows you to 
directly work with columns you've just created. 

```{r mutate}
diamonds %>%
  dplyr::select(color, carat, price) %>%
  mutate(ppc = price / carat, 
         ppc_rnd = round(ppc, 2))
```

<hr>

<b>Summarise values via `summarise`</b>

Similar to `plyr::summarise`, the **dplyr** version of `summarise` lets you melt 
a data frame into a single row depending on the supplied function. The function 
works quite similar as base-R `aggregate`. However, the ease of use is definitely 
on the side of `summarise`. 

<hr>

<b>Task: base-R data aggregation</b>

In order to demonstrate **dplyr**'s ease of use in terms of data summarization, 
try to `aggregate` diamonds in such a fashion that you end up with a data frame 
showing the minimum, mean and maximum price per diamond color.

<center>
  <img src="https://upload.wikimedia.org/wikipedia/commons/2/25/Hourglass_2.svg" alt="hourglass" style="width: 125px;"/>
</center>

```{r aggregate, echo = FALSE, eval = FALSE}
aggregate(diamonds$price, by = list(diamonds$color), FUN = function(x) {
  c(MIN = min(x), MEAN = mean(x), MAX = max(x))
})
```

As for the **dplyr** solution, the `group_by` function comes in handy when 
trying to calculate metrics from sub-groups of data, e.g. depending on the 
diamonds' color, rather than from the entire data frame.

```{r group_by}
diamonds %>% 
  group_by(color) %>%
  summarise(MIN = min(price), MEAN = mean(price), MAX = max(price))
```

In order to summarize an entire variable into one single value, just drop the 
`group_by` function. 

```{r summarise}
summarise(diamonds, min = min(price), mean = mean(price), max = max(price))
```

Now that you're familiar with the basics of data manipulation via **dplyr** and 
before moving on to actual data visualization via **ggplot2**, it's time to have 
a brief look at data format conversion from wide to long format which is essential 
for displaying numerous groups of data in one single plot.


## Data format conversions

Usually, your data is arranged in matrix format with rows and columns 
representing observations and variables, respectively. This is the common case 
not only for R, but also for many other statistical software packages, including 
Excel and SPSS. 
However, certain R packages, including visualization via **ggplot2**, are easier to handle when using a long data format rather than a wide one, thus requiring a "rearrangement of the form, but not the content, of the data" [@Wickham2007] according to so-called 
identifier (or ID) and measured variables. Here's a simple example taken from 
[Cookbook for R](http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/) 
to illustrate this point. Let's start off with the wide format:

```{r wide_format}
olddata_wide <- read.table(header = TRUE, text = '
 subject sex control cond1 cond2
       1   M     7.9  12.3  10.7
       2   F     6.3  10.6  11.1
       3   F     9.5  13.1  13.8
       4   M    11.5  13.4  12.9
')
```

Notice the strict structure of the `data.frame` with observations of subjects 
1-4 (two males and two females) arranged in rows and measured variables ('control', 
'cond1', 'cond2') in columns? Now, let's see what the long version of the 
dataset would look like:

```{r long_format}
library(reshape2)
melt(olddata_wide, id.vars = c("subject", "sex"))
```

What just happened? `melt` (or, more precisely, `melt.data.frame`) from the 
**reshape2** package was cast upon the dataset, forcing an increase in `nrow` to 
the expense of `ncol`. Columns 'subject' and 'sex' were thereby specified as 
ID variables, which prevented them from being split apart. The remaining variables, 
on the other side, now got arranged among each other rather than next to each other, 
with each row representing a unique ID-variable combination. 
You will notice later on that such a format is much easier to handle when it 
comes to visualizing grouped variables.

<hr>

<b>Task: Melting `diamonds`</b>

The 'diamonds' dataset included in **ggplot2** features quite a variety of 
variables per specimen (see `?diamonds` for further information). For the time 
being, let's assume we want to convert the dataset from wide into long format, 
thus resulting in a significant reduction of columns. Since no 'true' ID column 
is included so far, create a new column 'ID' from `rownames(diamonds)` and 
subsequently `melt` the dataset using all factor columns and the newly 
established ID column as `id.vars`.

<center>
  <img src="https://upload.wikimedia.org/wikipedia/commons/2/25/Hourglass_2.svg" alt="hourglass" style="width: 125px;"/>
</center>


## Visualizing data using **ggplot2**

Now that you've grown familiar with the basics of data manipulation using 
**dplyr** and **reshape2**, the next logical step in our workflow is to 
visualize the data that we've exhaustively prepared. Be aware that there are 
plenty of marvellous graphics packages available in R depending on which tasks 
you'd like to perform. Within the scope of our short course, **ggplot2** will do 
just fine not least because it is closely linked to the previously performed 
manipulation tasks (maybe because of the same author). 

```{r package}
## load 'ggplot2' package
library(ggplot2)
```

In **ggplot2**, every plot we want to draw consists of subsequent calls on how 
to represent the individual plot components (basically `x` and `y`). This means 
a much more consistent way of *building* visualizations as compared, for 
instance, to the much more conventional style of **lattice**, but 
it also means that things are rather different from what you might have learned 
about syntax and structure of (plotting) objects in R so far.

Note also that the brief examples shown below are mainly taken from another MARA course 
about [Creating publication quality graphs using R](https://oer.uni-marburg.de/data/mriliasmooc/lm_data/lm_1901/index.html). 
Feel free to browse the text tutorial or the accompanying [course slides](http://moc.environmentalinformatics-marburg.de/doku.php?id=courses:sp:pqgr) 
in case you're not entirely satisfied with the rather shallow introduction given 
here or if you wish to learn more about **lattice**-style figures. 

<hr>

<b>Scatter plots</b>

Scatter plots are produced by invoking `geom_points()`. If we wanted to plot 
variables 'price' and 'carat' from the 'diamonds' dataset against each other, 
the `ggplot` call would be as follows.

```{r gg-scat, tidy = FALSE, fig.cap = "A basic scatter plot created with ggplot2", fig.align = 'center'}
ggplot(aes(x = carat, y = price), data = diamonds) + 
  geom_point()
```

Let's look at the above code in a little more detail. The first line is the fundamental definition of **what** we want to plot. We provide the 'aesthetics' for the plot (`aes()`) and state that we want the values on the `x`-axis (`y`-axis) to represent 'carat' ('price'). The `data` to take these variables from is the 'diamonds' dataset. That's basically it, and this will not change a hell of a lot in the subsequent plotting routines.

What will change in the code chunks that follow is **how** we want the relationship between these variables to be represented in our plot. This is done by defining so-called geometries (`geom_...()`). In this first case we stated that we want the relationship between `x` and `y` to be rpresented as points, hence we used `geom_point()`.

For the sake of practicing, let's add another layer to our plot. In order to provide the regression line, for instance, we need a function called `stat_smooth()`. Note that we also specify the smoothing `method` and the line `colour`, `size` and `linetype`. `se = FALSE` tells the function not to display confidence intervals for the moment. See `?stat_smooth` for a more detailed overview of costumization possibilities.

```{r gg-scat-stat, tidy = FALSE, fig.cap = "A scatter plot incl. regression line.", fig.aline = 'center'}
ggplot(aes(x = carat, y = price), data = diamonds) + 
  geom_point() +
  stat_smooth(method = "lm", se = FALSE, 
              colour = "red", size = 1.5, linetype = "dashed")
```

If we wanted to provide a plot showing the relationship between 'price' and 'carat' in panels representing the quality of the diamonds, we need what in **ggplot2** is called facetting (similar to panels in **lattice**). To achive this, we simply repeat our plotting call from earlier and add another layer to the call which does the facetting. Note that this time, confidence intervals are included via `se = TRUE` and `fill`-ed grey.

```{r gg-scat-facet, tidy = FALSE, fig.cap = "ggplot2 version of a facetted plot", fig.align = 'center'}
ggplot(aes(x = carat, y = price), data = diamonds) + 
  geom_point() +
  stat_smooth(method = "lm", se = TRUE, 
              colour = "red", fill = "grey35") + 
  facet_wrap(~ cut)
```

One thing that some people dislike about the default settings in **ggplot2** is the grey background of the plots. This grey background is, in our opinion, a good idea when colors are involved as it tends to increase the contrast of the colors. If, however, the plot is a simple black-and-white scatter plot, a white facet background seems more reasonable. We can easily change this using a pre-defined theme called `theme_bw()`. Note that in the following, we also change the axis titles (`labs`) and the number of rows (`nrow`) and columns (`ncol`) into which the facets should be arranged - simple and straightforward.

```{r gg-scat-facet_bw, tidy = FALSE, fig.cap = "A ggplot2 panel plot with modified theme and added regression lines and confidence bands for each panel", fig.align = 'center'}
ggplot(aes(x = carat, y = price), data = diamonds) + 
  geom_point(colour = "grey65") +
  stat_smooth(method = "lm", se = TRUE, 
              colour = "black", fill = "grey35") +
  facet_wrap(~ cut, nrow = 2, ncol = 3) +
  labs(x = "Carat", y = "Price ($)") +
  theme_bw()
```

<hr>

<b>Box and whisker plots</b>

Box and whisker plots drawn with **ggplot2** look quite nice and, in contrast to **lattice**, do not require the user to exhaustively modify graphical parameter settings in order to get an acceptable result. But see for yourselves (using the default settings).

```{r gg-bw, tidy = FALSE, fig.cap = "A basic ggplot2 bw-plot", fig.align = 'center'}
## basic frame
p <- ggplot(aes(x = color, y = price), data = diamonds)

## add boxplot
p_bw <- p + geom_boxplot()

## print
print(p_bw)
```

Note that this time, we stored the basic frame of our plot in a variable `p` and added the desired `geom_boxplot()` layer afterwards. In order to keep a clear structure (and not to get in trouble with Tim), this is possibly the ideal way to create graphics using **ggplot2** - so stick to it! 

As we've already seen, the facetting is also just one more line of code.

```{r gg-bw-facet, tidy = FALSE, fig.cap = "ggplot2 panel bw-plot", fig.align = 'center'}
## add faceted boxplots
p_bw <- p + 
  geom_boxplot(fill = "grey90") +
  facet_wrap(~ cut) +
  theme_bw()

print(p_bw)
```

<hr>

<b>Histograms and density plots</b>

Much like with the box and whisker plot, **ggplot2** produces quite nice histograms and density plots when using the default settings.

```{r gg-hist, tidy = FALSE, fig.cap = "A basic histogram produced with ggplot2", fig.align = 'center', message=FALSE}
## new basic frame
p <- ggplot(aes(x = price), data = diamonds)

## add histogram
p_hist <- p + geom_histogram()

print(p_hist)
```

When working with **ggplot2** densityplots, it is quite easy to fill the area under the curve which really contributes to the visual representation of the data.

```{r gg-dens, tidy = FALSE, fig.cap = "A basic density plot produced with ggplot2", fig.align = 'center'}
## add densityplot
p_dens <- p +  geom_density(fill = "black", alpha = 0.5)

print(p_dens)
```

In addition to `theme_bw` which you already encountered before, there's a lot of different `theme` settings that let you customize your plots at will. For instance, let us slightly rotate the tick labels on the x-axis and move them horizontally a little bit. `element_text` is an essential function to achieve such things as it allows you to modify, amongst others, 

* font `family` and `face` (e.g. 'bold'),
* text `colour` and `size`,
* horizontal (`hjust`) and vertical (`vjust`) justification as well as 
* `angle` of the tick labels. 

(In case you prefer American English, `element_text` also understands `color` instead of `colour` &#9786;)

```{r gg-facet-density, tidy = FALSE, fig.cap = "A ggplot2 panel density plot conditioned according to 2 variables", fig.align = 'center'}
p_dens <- p +
  geom_density(fill = "black", alpha = 0.5) +
  facet_grid(color ~ cut) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_dens)
```
