```{r knitsetup, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=FALSE}
library(knitr)

opts_knit$set(base.dir='./', fig.path='', out.format='md')
opts_chunk$set(prompt=FALSE, comment='', results='markup')
# See yihui.name/knitr/options for more Knitr options.
##### Put other setup R code here


# end setup chunk
```
# Final remarks

We sincerely hope that this tutorial provides some useful parts to enhance your data analysis workflow. For a more general approach to software design (every function that you write is a software!) have a look at [this publication](http://www.plosbiology.org/article/fetchObject.action?uri=info:doi/10.1371/journal.pbio.1001745&representation=PDF).

This sources of this tutorial are hosted at [GitHub](https://github.com/marburg-open-courseware/advanced-programming-in-r)

Bug reports or feature requests should be filed at https://github.com/marburg-open-courseware/advanced-programming-in-r/issues

In case you want to provide feedback, don't hesitate to contact [Florian Detsch](http://umweltinformatik-marburg.de/mitarbeiterinnen-und-mitarbeiter/florian-detsch/) or 
[Tim Appelhans](http://umweltinformatik-marburg.de/mitarbeiterinnen-und-mitarbeiter/tim-appelhans/)


```{r mapview, echo = FALSE, eval = FALSE, fig.height = 2.5, fig.width = 9.47, warning = FALSE, message = FALSE}
library(jpeg)
library(rstudioapi)
print(getwd())
dir <- tempfile()
dir.create(dir)
img1 <- readJPEG("../../inst/portrait_header_fdetsch_01.jpg")
head(img1)
img2 <- readJPEG("../../inst/portrait_header_tappelhans_02.jpg")
fl1 <- paste0(dir, "/img1", ".png")
fl2 <- paste0(dir, "/img2", ".png")
png::writePNG(img1, fl1)
png::writePNG(img2, fl2)
htmlFile <- file.path(dir, "index.html")
pop <- paste(readLines("../../inst/slider.html"), collapse = "\n")
head(pop)
write(pop, htmlFile)
print(htmlFile)
cat(readLines(htmlFile), sep = "\n")
```

