```{r knitsetup, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=FALSE}
library(knitr)

opts_knit$set(base.dir='./', fig.path='', out.format='md')
opts_chunk$set(prompt=FALSE, comment='', results='markup')
# See yihui.name/knitr/options for more Knitr options.
##### Put other setup R code here


# end setup chunk
```
# Final remarks

We sincerely hope that this tutorial provides some useful parts to enhance your data analysis workflow. For a more general approach to software design (every function that you write is a software!) have a look at [this publication](http://www.plosbiology.org/article/fetchObject.action?uri=info:doi/10.1371/journal.pbio.1001745&representation=PDF).

This sources of this tutorial are hosted at [GitHub](https://github.com/marburg-open-courseware/advanced-programming-in-r)

Bug reports or feature requests should be filed at https://github.com/marburg-open-courseware/advanced-programming-in-r/issues

In case you want to provide feedback, don't hesitate to contact [Florian Detsch](http://umweltinformatik-marburg.de/mitarbeiterinnen-und-mitarbeiter/florian-detsch/) or 
[Tim Appelhans](http://umweltinformatik-marburg.de/mitarbeiterinnen-und-mitarbeiter/tim-appelhans/)


```{r mapview, echo = FALSE, eval = FALSE, fig.height = 2.5, fig.width = 9.47, warning = FALSE, message = FALSE}
library(jpeg)
library(rstudioapi)

dir <- tempfile()
dir.create(dir)
img1 <- readJPEG("../../inst/portrait_header_fdetsch_01.jpg")
img2 <- readJPEG("../../inst/portrait_header_tappelhans_02.jpg")
fl1 <- paste0(dir, "/img1", ".png")
fl2 <- paste0(dir, "/img2", ".png")
png::writePNG(img1, fl1)
png::writePNG(img2, fl2)
htmlFile <- file.path(dir, "index.html")
pop <- paste(readLines("../../inst/slider.html"), collapse = "\n")
write(pop, htmlFile)
```

<html>
<head>
<style>

/* image slider */
.photos {
    cursor: ew-resize;
    position:relative;
    overflow:visible;
    height: 100%;
    width: 100%;
    margin-top: 20px;
}
.photos, .photos .layer {
    height: 100%;
}

/* actual vertical image slider */
.photos .separator {
    border-left: 0px solid #000;
    border-right: 0px solid #000;
    height: 100%;
    left: -1px;
    line-height: 20px;
    overflow: visible;
    position: absolute;
    top: -20px;
    width: 1px;
    z-index: 2;
}
/* right slider name box */
.photos .separator div {
    background: rgba(97,78,153,0.6);
    background: #333;
    border-left: 1px solid #fff;
    color: #F2EFE9;
    font-family: 'Bree Serif', sans-serif;
    font-weight: bold;
    font-size: 12px;
    line-height: 20px;
    position: absolute;
    left: 0px;
    top: 0px;
    text-align: center;
    width: 100px;
    z-index: 9;
    border-radius: 5px 50px 50px 5px
}
/* left slider name box */
.photos .separator div.before {
    left: -100px;
    border-left: none;
    border-right: 1px solid #fff;
    border-radius: 50px 5px 5px 50px
}

#photo-before {
    max-width:100%;
    min-width:100%;
    position:absolute;
    background: url("http://umweltinformatik-marburg.de/uploads/pics/portrait_header_fdetsch_01.jpg") no-repeat;
    background-size:100% auto !important;
}
#photo-after {
    max-width:100%;
    min-width:100%;
    position:absolute;
    background: url("http://umweltinformatik-marburg.de/uploads/pics/portrait_header_tappelhans_02.jpg") no-repeat;
    background-size:100% auto !important;
}
</style>
</head>

<body>

<div class='photos'>
    <div class="separator">
        <div class="before">Flo</div>
        <div class="after">Tim</div>
    </div>
    <div class='layer' id='photo-after'></div>
    <div class='layer' id='photo-before'></div>
</div>


<script src='http://d3js.org/d3.v2.min.js'></script>
<script src='http://d3js.org/d3.v2.js'></script>

<script>
/* Update photo urls */
var updatePhotos = function(b,a) {
  if(b == 'custom' || a == 'custom') {
    b = document.getElementById('before_pic').value;
    a = document.getElementById('after_pic').value;
    if( !b.length || !a.length) {
      alert('Enter image URLs');
      return;
    }
  }
  document.getElementById('photo-before').style.background="url("+b+") no-repeat";
  document.getElementById('photo-after').style.background="url("+a+") no-repeat";
}


/* Gettin down to D3 business */
var beforeLayer = d3.select('#photo-before').node();

d3.select('.photos')
    .on('mousemove', function() {
        var pos = d3.mouse(this);
        d3.timer.flush();
        d3.timer(function() {
            d3.select('.separator').style("left", pos[0] + 'px');
            beforeLayer.style.clip = 'rect(0px ' + pos[0] + 'px 9999999px 0px)';
            return true;
        }, 0);
    });


</script>
</body>
</html>