```{r knitsetup, echo=FALSE, results='hide', warning=FALSE, message=FALSE, cache=FALSE}
library(knitr)

opts_knit$set(base.dir='./', fig.path='', out.format='md')
opts_chunk$set(prompt=TRUE, comment='', results='markup')
# See yihui.name/knitr/options for more Knitr options.
##### Put other setup R code here


# end setup chunk
```
# Final remarks

We sincerely hope that this tutorial provides some useful parts to enhance your data analysis workflow. For a more general approach to software design (every function that you write is a software!) have a look at [this publication](http://www.plosbiology.org/article/fetchObject.action?uri=info:doi/10.1371/journal.pbio.1001745&representation=PDF).

This sources of this tutorial are hosted at [GitHub](https://github.com/marburg-open-courseware/advanced-programming-in-r)

Bug reports or feature requests should be filed at https://github.com/marburg-open-courseware/advanced-programming-in-r/issues

In case you want to provide feedback, don't hesitate to contact [Florian Detsch](http://umweltinformatik-marburg.de/mitarbeiterinnen-und-mitarbeiter/florian-detsch/) or 
[Tim Appelhans](http://umweltinformatik-marburg.de/mitarbeiterinnen-und-mitarbeiter/tim-appelhans/)


```{r mapview, echo = FALSE, eval = FALSE, fig.height = 2.5, fig.width = 9.47, warning = FALSE, message = FALSE}
library(mapview)
library(jpeg)
library(raster)

### Flo
flo_image <- "http://umweltinformatik-marburg.de/uploads/pics/portrait_header_fdetsch_01.jpg"
flo_jpg <- readJPEG(readBin(flo_image, "raw", 1e6))

# Convert imagedata to raster
rst_blue <- raster(flo_jpg[, , 1])
rst_green <- raster(flo_jpg[, , 2])
rst_red <- raster(flo_jpg[, , 3])

flo <- brick(rst_red, rst_green, rst_blue)

### Tim
tim_image <- "http://umweltinformatik-marburg.de/uploads/pics/portrait_header_tappelhans_02.jpg"
tim_jpg <- readJPEG(readBin(tim_image, "raw", 1e6))

# Convert imagedata to raster
rst_blue <- raster(tim_jpg[, , 1])
rst_green <- raster(tim_jpg[, , 2])
rst_red <- raster(tim_jpg[, , 3])

tim <- brick(rst_red, rst_green, rst_blue)

m <- viewRGB(tim) + viewRGB(flo)
m@map
```
